\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-22 paper.tex
kyle's comments (sections I and II)

In this note, we develop a likelihood-free inference algorithm for fitting a
forward non-differentiable generative model to observed data. We propose to
adapt the adversarial training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the
adversarial objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an abitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

We consider a family of parameterized densities $p_\theta(\mathbf{x})$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\theta$ are the
parameters of interest. The simulation may involve some complicated latent
process, such that
\begin{equation}\label{eqn:p_x}
    p_\theta(\mathbf{x}) = \int p_\theta(\mathbf{x}|\mathbf{z}) p(\mathbf{z}) d\mathbf{z}
\end{equation}
where $\mathbf{z} \in \mathbb{R}^m$ is a latent variable providing an external source
of randomness.

We assume that we already have an accurate simulation of the stochastic
generative process that defines $p_\theta(\mathbf{x}|\mathbf{z})$, as
specified through a deterministic function $g(\cdot; \theta) : \mathbb{R}^m \to
\mathbb{R}^d$. That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p_\theta \equiv \mathbf{z} \sim p_z, \mathbf{x} = g(\mathbf{z}; \theta)
\end{equation}
such that the likelihood $p_\theta(\mathbf{x})$ can be rewritten as
\begin{equation}\label{eqn:p_x_sim}
    p_\theta(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p(\mathbf{z}) d\mathbf{z}.
\end{equation}
Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
used to generate data in forward mode. For this reason, evaluating the integral
in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\theta$,
e.g. as when $g$ is specified as a computer program.

@@

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p_\psi \equiv \mathbf{z} \sim p_z, \theta \sim q_\psi, \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q_\psi$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r$.
More formally, the learned model $p_\psi(\mathbf{x})$ therefore corresponds to the marginal distribution
$\int q_\psi(\theta) p_\theta(\mathbf{x}) d\theta$ of the generated data.

>>>>

In this note, we develop a likelihood-free inference algorithm for the point
estimation from observed data of the parameters of a forward non-differentiable
generative model. We propose to adapt the adversarial
training procedure of generative adversarial
networks~\cite{goodfellow2014generative} by replacing the implicit generative
network with a domain-based scientific simulator, and solve the resulting
non-differentiable minimax problem by minimizing variational upper
bounds~\citep{2011arXiv1106.4487W,2012arXiv1212.4507S} of the adversarial
objectives. The procedure results in learning a proposal distribution over
simulator parameters, hence producing an abitrarily tight family of models whose
joint collection of generated samples matches the observed data.

@@

We consider a family of parameterized densities $p_\theta(\mathbf{x})$
defined implicitly through the simulation of a stochastic generative process,
where $\mathbf{x} \in \mathbb{R}^d$ is the data and $\theta$ are the
parameters of interest. The simulation may involve some complicated latent
process, such that
\begin{equation}\label{eqn:p_x}
    p_\theta(\mathbf{x}) = \int p_\theta(\mathbf{x}|\mathbf{z}) p_{z(\theta)}(\mathbf{z}) d\mathbf{z}
\end{equation}
where $\mathbf{z} \in {\cal Z}$ is a latent variable providing an external
source of randomness. In particular, $\mathbf{z}$ is not necessarily assumed to
be a fixed-size vector (e.g., it can be a sequence of variable length) and its
distribution $p_{z(\theta)}$ may itself depend on $\theta$ in some intricated way.

We assume that we already have an accurate simulation of the stochastic
generative process that defines $p_\theta(\mathbf{x}|\mathbf{z})$, as
specified through a highly regularized deterministic function $g(\cdot; \theta) : {\cal Z} \to
\mathbb{R}^d$ with usually few parameters. That is, we consider
\begin{equation}\label{eqn:p_theta}
    \mathbf{x} \sim p_\theta \equiv \mathbf{z} \sim p_{z(\theta)}, \mathbf{x} = g(\mathbf{z}; \theta)
\end{equation}
such that the likelihood $p_\theta(\mathbf{x})$ can be rewritten as
\begin{equation}\label{eqn:p_x_sim}
    p_\theta(\mathbf{x}) = \frac{\partial}{\partial x_1} \dots \frac{\partial}{\partial x_d} \int_{\{\mathbf{z}:g(\mathbf{z};\theta) \leq \mathbf{x}\}} p_{z(\theta)}(\mathbf{z}) d\mathbf{z}.
\end{equation}
\glnote{Is it correct to write $d\mathbf{z}$ given the undefined structure of ${\cal Z}$?}
Importantly, the simulator $g$ is assumed to be a non-invertible function, that can only be
used to generate data in forward mode. For this reason, evaluating the integral
in Eqn.~\ref{eqn:p_x_sim} is intractable. As commonly found
in science, we finally assume the lack of access to or existence of derivatives of $g$ with respect to $\theta$,
e.g. as when $g$ is specified as a computer program.

@@

Practically, the variational objectives \ref{eqn:vo-ud}-\ref{eqn:vo-ug}
have the effect of replacing the modeled data distribution of Eqn.~\ref{eqn:p_theta} with
a distribution parameterized in terms of $\psi$:
\begin{equation}\label{eqn:p_psi}
    \mathbf{x} \sim p_\psi \equiv \mathbf{z} \sim p_z, \theta \sim q_\psi, \mathbf{x} = g(\mathbf{z}; \theta).
\end{equation}
Intuitively, this corresponds to a family of simulators, each configured
with randomly sampled parameters $\theta \sim q_\psi$, whose joint collection
of generated samples is optimized with adversarial training to approach the real data distribution $p_r$.
More formally, the learned model $p_\psi(\mathbf{x})$ therefore corresponds to the marginal distribution
$\int q_\psi(\theta) p_\theta(\mathbf{x}) d\theta$ of the generated data.
\glnote{Connection with variational bayes?}
